---
title: "Usefull phenotypical labels for the Subjects in the PsyCourse data"
output: html_notebook
---
```{r SETUP Load packages}
#library(ranger)
#library(caret)
library(tidyverse)
#library(ggplot2)
```

```{r SETUP Global chunk options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.height = 3.5,
                      fig.width = 4)
```

```{r SETUP Clean-up, eval=T}
# remove all objects to start with clean environment
rm(list=ls())
```

```{r SETUP Functions}
check_two_true_in_a_row <- function(vector) {
  any(vector & lead(vector, default = FALSE))
}
```

```{r DATA Loading}
load("231107_PsyCourse6.0_Pheno_Proposal_69_Schwarz.RData")
data <- prop
rm(prop)
```

```{r DATA Create diagnosis filter vectors}
## Captures both the ones labled "Schizophrenia" (DSM 295.10/20/30/60/90) and "ICD-10 Schizophrenia" (for some participants the ICD diagnosis was not reassessed with the DSM framework)
F20x_id <-
data %>%
  filter(v1_stat == "CLINICAL") %>% 
  filter(str_detect(v1_scid_dsm_dx_cat, "Schizophrenia")) %>% 
  pull(v1_id)

# Captures all patients labeld as "Schizoaffective Disorder" (DSM 295.70)
F25x_id <-
  data %>%
  filter(v1_stat == "CLINICAL") %>% 
  filter(str_detect(v1_scid_dsm_dx_cat, "Schizoaffective Disorder")) %>% 
  pull(v1_id)

# maybe mapping to ICD not quite correct
F23x_id <-
data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, v1_scid_dsm_dx_cat) %>% 
  filter(str_detect(v1_scid_dsm_dx_cat, "Schizophreniform Disorder | Brief Psychotic Disorder")) %>% 
  pull(v1_id)

# Captures both "Bipolar-I Disorder" (DSM 296.X) and Bipolar-II Disorder (DSM 296.89)
F31x_id <-
  data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, v1_scid_dsm_dx_cat) %>% 
  filter(str_detect(v1_scid_dsm_dx_cat, "Bipolar")) %>% 
  pull(v1_id)

# Lists all the distinct diagnosis in this specific version fo the PsyCourse data 
data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_scid_dsm_dx_cat) %>% 
  distinct()
```

# Introduction
In this file different labels or new dimensions for the subjects in the PsyCourse datafile are created based on the phenotypical data.
The PsyCourse study is described here: http://www.psycourse.de/index.html and in this study http://dx.doi.org/10.1002/ajmg.b.32639
The codebook decribing all the variables in the acquired dataset can be found here: https://data.ub.uni-muenchen.de/251/1/210908_PsyCourse_v5.0.html

Depending on the complexity of the label there are some characteristics of the dataset which complicate things. Some of these are:

* Missings
  + In general the most data is usually available at the first timepoint, getting sparse over the time course
  + The missings are usually not random, with e.g. more severe cases are more likely to drop out after the first timepoint

# Phenotypical Labels
In the PsyCourse study the subjects were measured at 4 different timepoints, each 6 months apart.
Creating phenotypical labels basically means dividing the subjects into meaningfull groups. These labels can be characterized on two basic dimensions:

* Crosssectional vs. longitudinal
* Based on domain knowledge vs data driven

Note that by the choice of the phenotypical measurements there is already a lot of domain knowledge genuinely in the dataset.

New dimensions are generally created based on crosssectional data.

## Crossectional

### DIMENSIONS PCA dimensions within PANSS (psychoticism, disorganization, negative symptoms)
TODO: So far the PCA of the PsyCourse PANSS data of timepoint 1 does not show the PCs reported in the literature. Find reasons.
#### Schizophrenic patients
The PANSS data of the schizophrenic patients at timepoint 1 is selected.
```{r DATA}
t1_schizo_panss_data <-
data %>%
  filter(v1_stat == "CLINICAL") %>%
  filter(v1_id %in% F20x_id) %>% 
  select(v1_id, matches("v1_panss_\\w\\d")) %>% 
  mutate_at(vars(matches("v1_panss")), as.numeric) %>% 
  na.omit()
```

PCA analysis on the data is performed.
```{r DATA ANA PCA inspectiony}
# Run PCA
pca_result <- prcomp(t1_schizo_panss_data %>% 
                       select(matches("v1_panss_\\w\\d")),
                     #select(matches("v1_panss_[p,n]\\d")),
                     scale = T
                     #,rank.=2
                     )

# Extract loadings for the first three principal components
loadings <- pca_result$rotation[, 1:3]

## make a scree plot
pca.var <- pca_result$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
 
barplot(pca.var.per, main="Scree Plot", xlab="Principal Component", ylab="Percent Variation")

# Plot the first two components against each other
pca.data <- data.frame(
  X=pca_result$x[,1],
  Y=pca_result$x[,2])
 
ggplot(data=pca.data, aes(x=X, y=Y)) +
  geom_point() +
  xlab(paste("PC1 - ", pca.var.per[1], "%", sep="")) +
  ylab(paste("PC2 - ", pca.var.per[2], "%", sep="")) +
  theme_bw() +
  ggtitle("PC1 vs PC2")

# Create a dataframe to store the variables and their loadings
loadings_df <- data.frame(
  Variable = colnames(t1_schizo_panss_data %>% 
                       select(matches("v1_panss_\\w\\d"))),
                     #select(matches("v1_panss_[p,n]\\d"))),
  PC1 = loadings[, 1],
  PC2 = loadings[, 2],
  PC3 = loadings[, 3]
)

# Identify variables with the highest loadings for each component
top_loadings_PC1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE), ][1:10, ]
top_loadings_PC2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE), ][1:10, ]
top_loadings_PC3 <- loadings_df[order(abs(loadings_df$PC3), decreasing = TRUE), ][1:10, ]

top_loadings_PC1
top_loadings_PC2
top_loadings_PC3


```


### LABEL Metabolic Syndrom
According to the criteria in http://dx.doi.org/10.1111/j.1464-5491.2006.01858.x (see table 5 in the paper for the worldwide definition) subjects with a metabolic syndrome are defined based on the somatic data from visit 1.

#### Limitations
Not all criteria for the metabolic syndrome are available in the data:

* Central obesity
 + waist circumference criteria is not ethnicity specific
* Raised triglycerides
 + only available in a dichotomous variable for triglyceride and cholesterol combined
* Reduced HDL-cholesterol
 + not available
* Raised blood pressure
 + only available as a dichotomous variable for systolic and diastolic combined
* Raised fasting plasma glucose
 + only criteria 'diabetes' available
 + no fasting plasma glucose or oral glucose tolerance test
 
#### Label
Part of the criteria is whether a person takes medication to treat hypertriglyceridamia or hypertension. These are criteria is not considered in this simple version of the label.

```{r DATA ANALYSIS All subjects metabolic syndrome simple at v1}
label_metabolic_syndrome_simple <-
  data %>% 
  select(v1_id, v1_sex, v1_bmi, v1_waist, v1_chol_trig, v1_hyperten, v1_diabetes) %>%
  rowwise() %>% 
  mutate(Two_Criteria = sum(c_across(v1_chol_trig:v1_diabetes) == "Y") >= 2) %>%
  ungroup() %>% 
  mutate(metabolic_syndrome = ifelse(((v1_sex == 'F' & v1_waist >= 80) | (v1_sex == "M" & v1_waist >=94) | (v1_bmi > 30)) & (Two_Criteria), TRUE, FALSE)) %>% 
  select(v1_id, metabolic_syndrome)
```

Read out of the categories and subcategories of the medications in the dataset.
```{r DATA ANALYSIS distinct categories of medications}
med_categories <-
v1_med_clin_orig %>% 
  select(starts_with ("v1_medikabehand3_med_kategorie_199998_")) %>% 
  pivot_longer(cols=everything(),
               names_to = "Variable",
               values_to = "Value") %>% 
  distinct(Value)  

med_subcategories <-
v1_med_clin_orig %>% 
  select(starts_with ("v1_medikabehand3_med_kategorie_sub_199998_")) %>% 
  pivot_longer(cols=everything(),
               names_to = "Variable",
               values_to = "Value") %>% 
  distinct(Value)
```

For this version of the label the  medication criteria is taken into account. 
```{r DATA ANALYSIS All subjects metabolic syndrome at v1}
label_metabolic_syndrome <-
data %>%
  left_join(x = .,
            y = v1_med_clin_orig,
            by = "v1_id") %>%
  mutate(central_obesity = if_else((v1_sex == 'F' & v1_waist >= 80) | (v1_sex == "M" & v1_waist >= 94) | (v1_bmi > 30), TRUE, FALSE, missing=FALSE),
         raised_fasting_plasma_glucose = v1_diabetes == 'Y'
  )%>% 
  rowwise() %>% 
  mutate(raised_triglycerides = if_else(v1_chol_trig == 'Y' | rowSums(across(starts_with("v1_medikabehand3_med_kategorie_sub_199998_"), ~ .x == "(Cholesetrinsenker)"), na.rm =TRUE)>0, TRUE, FALSE),
         raised_bloodpressure = ifelse(v1_hyperten == 'Y' | rowSums(across(starts_with("v1_medikabehand3_med_kategorie_sub_199998_"), ~ .x == "Antihypertensiva"), na.rm = TRUE)>0, TRUE, FALSE))%>%
  ungroup()%>%
  mutate(metabolic_syndrome = central_obesity &
           ((raised_triglycerides + raised_bloodpressure + raised_fasting_plasma_glucose) >= 2)) %>% 
  select(v1_id, metabolic_syndrome, central_obesity)
```

This overview shows that, as expected, there is not much of a difference between the full and the simple version of the label.
```{r DATA overview over the results}
summary(label_metabolic_syndrome)
summary(label_metabolic_syndrome_simple)
```

Some basic comparisons showing examples how the label could be used.
```{r DATA ANA Some simple analyses to get an idea, eval=FALSE}
# ~twice as likely to have a metabolic syndrome as a patient compared to a HC
data %>% 
  select(v1_id, v1_stat) %>% 
  left_join(x = .,
            y = label_metabolic_syndrome,
            by = "v1_id") %>% 
  select(v1_stat, metabolic_syndrome) %>% 
  table() %>% 
  fisher.test()

# ~twice as likely to have central obesity as a patient
data %>% 
  select(v1_id, v1_stat) %>% 
  left_join(x = .,
            y = label_metabolic_syndrome,
            by = "v1_id") %>% 
  select(v1_stat, central_obesity) %>% 
  table() %>% 
  fisher.test()

# ~twice as likely to have a metabolic syndrome if you ever took lithium as a patient
data %>%
  filter(v1_stat=="CLINICAL", v1_lith != -999) %>% 
  select(v1_id, v1_lith) %>%
  droplevels() %>% 
  left_join(x = .,
            y = label_metabolic_syndrome,
            by = "v1_id") %>% 
  select(v1_lith, metabolic_syndrome) %>%
  table() %>% 
  fisher.test()

# >twice as likely to have metabolic syndrome when taking any antidepressant medication as a patient 
data %>% 
  filter(v1_stat == "CLINICAL") %>% 
  select(v1_id, v1_Antidepressants) %>% 
  mutate(v1_antidep_ever = ifelse(v1_Antidepressants > 0, TRUE, FALSE)) %>% 
  droplevels() %>% 
  left_join(x = .,
            y = label_metabolic_syndrome,
            by = "v1_id") %>% 
  select(v1_antidep_ever, metabolic_syndrome) %>%
  table() %>% 
  fisher.test()

```

## Longitudinal

### LABEL Remission within the 18 months time course (never/ever/always)

#### Schizophrenic patients

* Only for patients with a diagnosis of "Schizophrenia (ICD F20.x)".
* Definition of remission based on: [url]http://dx.doi.org/10.1176/appi.ajp.162.3.441
  + The criteria based on the PANSS are:
  ++ Items P1, P2, P3, N1, N4, N6, G5, G9 =< 3
  ++ For >= 6 months

##### Missings  
```{r DATA ANALYSIS Table of missing pattern PANSS F20x remission criteria items}
# The 8 items from the PANSS defining remission are saved in a vector
search_vector  = c("p1","p2","p3","n1","n4","n6","g5","g9")

data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, contains(search_vector)) %>%
  rename(ID = v1_id) %>%
  filter(ID %in% F20x_id) %>%
  mutate_at(vars(matches("v(\\d+)_panss")), as.numeric) %>% 
  pivot_longer(cols = -ID,
               names_to = c("time_point", "item"),
               names_pattern = "v(\\d+)_panss_(.+)",
               values_to = "value") %>%
  group_by(ID, time_point) %>% 
  summarise(miss = ifelse(any(is.na(value)), '-', '+')) %>% 
  summarise(pattern = toString(miss)) %>% 
  count(pattern) %>%
  mutate(percentage = round((n/sum(n)*100), 2)) %>% 
  arrange(desc(n))
```

##### Label
At the moment only implemented for the cases with complete data on all 4 timepoints.
TODO: Implement ever/never using all cases which have complete data for at least 2 timepoints in a row 
```{r DATA ANALYSIS F20x fullremission ever}
search_vector  = c("p1","p2","p3","n1","n4","n6","g5","g9")

label_remission <-  
  data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, contains(search_vector)) %>%
  filter(v1_id %in% F20x_id) %>% 
  mutate_at(vars(matches("v(\\d+)_panss")), as.numeric) %>%
  na.omit() %>% 
  pivot_longer(cols = -v1_id,
               names_to = c("time_point", "item"),
               names_pattern = "v(\\d+)_panss_(.+)",
               values_to = "value") %>% 
  group_by(v1_id, time_point) %>% 
  summarise(remission=all(value<=3)) %>%
  left_join(x=.,
            y= data %>%
              select(v1_id, contains("clin_ill_ep_snc_lst")) %>%
              rename(v1_id = v1_id) %>%
              filter(v1_id %in% F20x_id) %>%
              mutate_at(vars(contains("clin_ill_ep_snc_lst")), ~if_else(.=="N",TRUE,FALSE, missing = TRUE)) %>% 
              pivot_longer(cols = -v1_id,
                           names_to = c("time_point"),
                           names_pattern = "v(\\d+)",
                           values_to = "in_remission_since_last_visit"),
            by = join_by(v1_id, time_point)
  ) %>% 
  group_by(v1_id) %>% 
  mutate(full_remission = as.logical(remission*lead(in_remission_since_last_visit, default = TRUE))) %>% 
  summarise(ever_in_full_remission = check_two_true_in_a_row(full_remission),
            always_in_full_remission = all(full_remission)) %>%
  ungroup() %>% 
  mutate(remission = factor(case_when(
    ever_in_full_remission & !always_in_full_remission ~ "ever",
    ever_in_full_remission & always_in_full_remission ~ "always",
    TRUE ~ "never"
  ))) %>% 
  select(v1_id, remission)
```

#### Bipolar patients
TODO: Finish the part with depression and adapt the code accordingly.

Remission in bipolar patients basically means to be in remission regarding mania and regarding depression.  

To define the remission regarding mania usually the Young Mania Rating Scale (YMRS) is used. But there are two different cutoffs being used:

* YMRS <= 12 (http://dx.doi.org/10.1097/chi.0b013e3180335ae4)
* YMRS <= 7 (http://dx.doi.org/10.1034/j.1399-5618.2003.02237.x, http://dx.doi.org/10.1016/j.jad.2007.06.011 - even argue for <= 4 but also support <=7)

To define the remission regarding depression in bipolar patients usually the MADRS and/or HAMD are used.
In the PsyCourse data the Inventory of depressive symptomatology (IDS-C30) and the BDI-II are available.
##### Missings
```{r DATA ANALYSIS Table of missing pattern YMRS sum_tot also F31x remission criterium}
data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, contains("ymrs_sum")) %>% 
  rename(ID = v1_id) %>%
  filter(ID %in% F31x_id) %>%
  pivot_longer(cols = starts_with("v"),
               names_to = "Time_Point",
               names_pattern = "v(\\d+)_ymrs_sum",
               values_to = "ymrs_sum_tot") %>%
  mutate(ymrs_sum_tot = ifelse(is.na(ymrs_sum_tot), "-", "+")) %>% 
  group_by(ID) %>% 
  summarise(pattern = toString(ymrs_sum_tot)) %>% 
  count(pattern) %>%
  mutate(percentage = round((n/sum(n)*100), 2)) %>% 
  arrange(desc(n))
```

##### Label
TODO: Add the control for the episode since last visit
```{r DATA ANALYSIS F31x remission ever}
#could not find a time criteria for the mania remission according to YMRS
Label_F31x_id_remission_ever <-
data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, contains("ymrs_sum")) %>%
  filter(v1_id %in% F31x_id) %>%
  pivot_longer(cols = -v1_id,
               names_to = c("time_point"),
               names_pattern = "v(\\d+)",
               values_to = "ymrs_sum_score") %>% 
  group_by(ID) %>% 
  summarise(ever_in_remission=any(ymrs_sum_score<=7, na.rm = T)) %>%
  select(v1_id, ever_in_remission)
```

### LABEL Weightgain under AP medication

#### Missings
```{r DATA ANALYSIS Table of missing pattern weight}
data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, contains("weight")) %>%
  mutate_at(vars(matches("v(\\d+)_weight")), as.numeric) %>% 
  pivot_longer(cols = -v1_id,
               names_to = c("time_point"),
               names_pattern = "v(\\d+)_weight",
               values_to = "value") %>%
  group_by(v1_id, time_point) %>% 
  summarise(miss = ifelse(any(is.na(value)), '-', '+')) %>% 
  summarise(pattern = toString(miss)) %>% 
  count(pattern) %>%
  mutate(percentage = round((n/sum(n)*100), 2)) %>% 
  arrange(desc(n))
```

```{r DATA LOADING Medication}
load("data/230614_v6.0_psycourse_clin_raw_med_visit1.RData")
load("data/230614_v6.0_psycourse_clin_raw_med_visit2.RData")
load("data/230614_v6.0_psycourse_clin_raw_med_visit3.RData")
load("data/230614_v6.0_psycourse_clin_raw_med_visit4.RData")
```

```{r DATA SELECTION}
data.weight.longi <-
  data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, contains("weight"))

#data.metabolicS.v1 <-
  


```



