---
title: "Usefull phenotypical labels for the Subjects in the PsyCourse data"
output: html_notebook
---
```{r SETUP Load packages}
#library(ranger)
#library(caret)
library(tidyverse)
#library(ggplot2)
```

```{r SETUP Global chunk options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.height = 3.5,
                      fig.width = 4)
```

```{r SETUP Clean-up, eval=T}
# remove all objects to start with clean environment
rm(list=ls())
```

```{r SETUP Functions}
check_two_true_in_a_row <- function(vector) {
  any(vector & lead(vector, default = FALSE))
}
```

```{r DATA Loading}
load("231107_PsyCourse6.0_Pheno_Proposal_69_Schwarz.RData")
data <- prop
rm(prop)
```

```{r DATA Create diagnosis filter vectors}
## Captures both the ones labled "Schizophrenia" (DSM 295.10/20/30/60/90) and "ICD-10 Schizophrenia" (for some participants the ICD diagnosis was not reassessed with the DSM framework)
F20x_id <-
data %>%
  filter(v1_stat == "CLINICAL") %>% 
  filter(str_detect(v1_scid_dsm_dx_cat, "Schizophrenia")) %>% 
  pull(v1_id)

# Captures all patients labeld as "Schizoaffective Disorder" (DSM 295.70)
F25x_id <-
  data %>%
  filter(v1_stat == "CLINICAL") %>% 
  filter(str_detect(v1_scid_dsm_dx_cat, "Schizoaffective Disorder")) %>% 
  pull(v1_id)

# maybe mapping to ICD not quite correct
F23x_id <-
data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, v1_scid_dsm_dx_cat) %>% 
  filter(str_detect(v1_scid_dsm_dx_cat, "Schizophreniform Disorder | Brief Psychotic Disorder")) %>% 
  pull(v1_id)

# Captures both "Bipolar-I Disorder" (DSM 296.X) and Bipolar-II Disorder (DSM 296.89)
F31x_id <-
  data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, v1_scid_dsm_dx_cat) %>% 
  filter(str_detect(v1_scid_dsm_dx_cat, "Bipolar")) %>% 
  pull(v1_id)

# Lists all the distinct diagnosis in this specific version fo the PsyCourse data 
data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_scid_dsm_dx_cat) %>% 
  distinct()
```

# Introduction
In this file different labels or new dimensions for the subjects in the PsyCourse datafile are created based on the phenotypical data.
The PsyCourse study is described here: [url]http://www.psycourse.de/index.html and in this study [url]http://dx.doi.org/10.1002/ajmg.b.32639
The codebook decribing all the variables in the acquired dataset can be found here: [url]https://data.ub.uni-muenchen.de/251/1/210908_PsyCourse_v5.0.html

Depending on the complexity of the label there are some characteristics of the dataset which complicate things. Some of these are:

* Missings
  + In general the most data is usually available at the first timepoint, getting sparse over the time course
  + The missings are usually not random, with e.g. more severe cases are more likely to drop out after the first timepoint

# Phenotypical Labels
In the PsyCourse study the subjects were measured at 4 different timepoints, each 6 months apart.
Creating phenotypical labels basically means dividing the subjects into meaningfull groups. These labels can be characterized on two basic dimensions:

* Crosssectional vs. longitudinal
* Based on domain knowledge vs data driven

Note that by the choice of the phenotypical measurements there is already a lot of domain knowledge genuinely in the dataset.

New dimensions are generally created based on crosssectional data.

## Crossectional

### DIMENSIONS PCA dimensions within PANSS (psychoticism, disorganization, negative symptoms)

#### Schizophrenic patients


```{r}
t1_schizo_panss_data <-
data %>%
  filter(v1_stat == "CLINICAL") %>%
  filter(v1_id %in% F20x_id) %>% 
  select(v1_id, matches("v1_panss_\\w\\d")) %>% 
  mutate_at(vars(matches("v1_panss")), as.numeric) %>% 
  na.omit()
```

```{r}
# Run PCA
pca_result <- prcomp(t1_schizo_panss_data %>% 
                       select(matches("v1_panss_\\w\\d")),
                     #select(matches("v1_panss_[p,n]\\d")),
                     scale = T
                     #,rank.=2
                     )

# Extract loadings for the first three principal components
loadings <- pca_result$rotation[, 1:3]

## make a scree plot
pca.var <- pca_result$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
 
barplot(pca.var.per, main="Scree Plot", xlab="Principal Component", ylab="Percent Variation")

# Plot the first two components against each other
pca.data <- data.frame(
  X=pca_result$x[,1],
  Y=pca_result$x[,2])
 
ggplot(data=pca.data, aes(x=X, y=Y)) +
  geom_point() +
  xlab(paste("PC1 - ", pca.var.per[1], "%", sep="")) +
  ylab(paste("PC2 - ", pca.var.per[2], "%", sep="")) +
  theme_bw() +
  ggtitle("PC1 vs PC2")

# Create a dataframe to store the variables and their loadings
loadings_df <- data.frame(
  Variable = colnames(t1_schizo_panss_data %>% 
                       select(matches("v1_panss_\\w\\d"))),
                     #select(matches("v1_panss_[p,n]\\d"))),
  PC1 = loadings[, 1],
  PC2 = loadings[, 2],
  PC3 = loadings[, 3]
)

# Identify variables with the highest loadings for each component
top_loadings_PC1 <- loadings_df[order(abs(loadings_df$PC1), decreasing = TRUE), ][1:10, ]
top_loadings_PC2 <- loadings_df[order(abs(loadings_df$PC2), decreasing = TRUE), ][1:10, ]
top_loadings_PC3 <- loadings_df[order(abs(loadings_df$PC3), decreasing = TRUE), ][1:10, ]

top_loadings_PC1
top_loadings_PC2
top_loadings_PC3


```



## Longitudinal

### LABEL Remission within the 18 months time course (never/ever/always)

#### Schizophrenic patients

* Only for patients with a diagnosis of "Schizophrenia (ICD F20.x)".
* Definition of remission based on: [url]http://dx.doi.org/10.1176/appi.ajp.162.3.441
  + The criteria based on the PANSS are:
  ++ Items P1, P2, P3, N1, N4, N6, G5, G9 =< 3
  ++ For >= 6 months

##### Missings  
```{r DATA ANALYSIS Table of missing pattern PANSS F20x remission criteria items}
# The 8 items from the PANSS defining remission are saved in a vector
search_vector  = c("p1","p2","p3","n1","n4","n6","g5","g9")

data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, contains(search_vector)) %>%
  rename(ID = v1_id) %>%
  filter(ID %in% F20x_id) %>%
  mutate_at(vars(matches("v(\\d+)_panss")), as.numeric) %>% 
  pivot_longer(cols = -ID,
               names_to = c("time_point", "item"),
               names_pattern = "v(\\d+)_panss_(.+)",
               values_to = "value") %>%
  group_by(ID, time_point) %>% 
  summarise(miss = ifelse(any(is.na(value)), '-', '+')) %>% 
  summarise(pattern = toString(miss)) %>% 
  count(pattern) %>%
  mutate(percentage = round((n/sum(n)*100), 2)) %>% 
  arrange(desc(n))
```

##### Label
At the moment only implemented for the cases with complete data on all 4 timepoints.
TODO: Implement ever/never using all cases which have complete data for at least 2 timepoints in a row 
```{r DATA ANALYSIS F20x fullremission ever}
search_vector  = c("p1","p2","p3","n1","n4","n6","g5","g9")

label_remission <-  
  data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, contains(search_vector)) %>%
  filter(v1_id %in% F20x_id) %>% 
  mutate_at(vars(matches("v(\\d+)_panss")), as.numeric) %>%
  na.omit() %>% 
  pivot_longer(cols = -v1_id,
               names_to = c("time_point", "item"),
               names_pattern = "v(\\d+)_panss_(.+)",
               values_to = "value") %>% 
  group_by(v1_id, time_point) %>% 
  summarise(remission=all(value<=3)) %>%
  left_join(x=.,
            y= data %>%
              select(v1_id, contains("clin_ill_ep_snc_lst")) %>%
              rename(v1_id = v1_id) %>%
              filter(v1_id %in% F20x_id) %>%
              mutate_at(vars(contains("clin_ill_ep_snc_lst")), ~if_else(.=="N",TRUE,FALSE, missing = TRUE)) %>% 
              pivot_longer(cols = -v1_id,
                           names_to = c("time_point"),
                           names_pattern = "v(\\d+)",
                           values_to = "in_remission_since_last_visit"),
            by = join_by(v1_id, time_point)
  ) %>% 
  group_by(v1_id) %>% 
  mutate(full_remission = as.logical(remission*lead(in_remission_since_last_visit, default = TRUE))) %>% 
  summarise(ever_in_full_remission = check_two_true_in_a_row(full_remission),
            always_in_full_remission = all(full_remission)) %>%
  ungroup() %>% 
  mutate(remission = factor(case_when(
    ever_in_full_remission & !always_in_full_remission ~ "ever",
    ever_in_full_remission & always_in_full_remission ~ "always",
    TRUE ~ "never"
  ))) %>% 
  select(v1_id, remission)
```

#### Bipolar patients
TODO: Fully explanation with literature YMRStot =< 7 or =< 12 etc

##### Missings
```{r DATA ANALYSIS Table of missing pattern YMRS sum_tot also F31x remission criterium}
data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, contains("ymrs_sum")) %>% 
  rename(ID = v1_id) %>%
  filter(ID %in% F31x_id) %>%
  pivot_longer(cols = starts_with("v"),
               names_to = "Time_Point",
               names_pattern = "v(\\d+)_ymrs_sum",
               values_to = "ymrs_sum_tot") %>%
  mutate(ymrs_sum_tot = ifelse(is.na(ymrs_sum_tot), "-", "+")) %>% 
  group_by(ID) %>% 
  summarise(pattern = toString(ymrs_sum_tot)) %>% 
  count(pattern) %>%
  mutate(percentage = round((n/sum(n)*100), 2)) %>% 
  arrange(desc(n))
```

##### Label
TODO: Add the control for the episode since last visit
```{r DATA ANALYSIS F31x remission ever}
#could not find a time criteria for the mania remission according to YMRS
F31x_id_remission_ever <-
data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, contains("ymrs_sum")) %>%
  rename(ID = v1_id) %>%
  filter(ID %in% F31x_id) %>%
  pivot_longer(cols = -ID,
               names_to = c("time_point"),
               names_pattern = "v(\\d+)",
               values_to = "ymrs_sum_score") %>% 
  group_by(ID) %>% 
  summarise(ever_in_remission=any(ymrs_sum_score<=7, na.rm = T)) %>%
  filter(ever_in_remission == T) %>% 
  pull(ID)

data %>%
  filter(v1_stat == "CLINICAL") %>%
  select(v1_id, contains("ymrs_sum")) %>%
  rename(ID = v1_id) %>%
  filter(ID %in% F31x_id) %>%
  pivot_longer(cols = -ID,
               names_to = c("time_point"),
               names_pattern = "v(\\d+)",
               values_to = "ymrs_sum_score") %>% 
  group_by(ID) %>% 
  summarise(ever_in_remission=any(ymrs_sum_score<=7, na.rm = T)) %>% 
  count(ever_in_remission)
```

### LABEL Weightgain under AP medication

```{r}

```

